### 2021-09-03 revision/ch5langevin.tex
added phd thesis.

High-speed video experiments have measured different streamwise particle velocity distributions without indicating why one distribution or another appears in a given set of hydraulic and sedimentary conditions.
One set of studies has shown exponential particle velocity distributions \citep{Charru2004,Lajeunesse2010,Roseberry2012,Seizilles2014,Fathel2015,Fathel2016}.
These experiments involve uniformly-sized small sands or glass beads ($~0.05-2$mm) in turbulent and subcritical flows, but not always; the experiments of \citet{Lajeunesse2010} were supercritical, while the experiments of \citet{Charru2004} and \citet{Seizilles2014} were viscous.
A second set of studies have shown Gaussian particle velocity distributions \citep{Ancey2014,Heyman2016,Martin2012}. In these experiments, particles are typically larger ($2-8$mm) uniformly-sized gravels or glass beads, and flows are generally turbulent and super-critical.
Other experiments display velocity distributions that are intermediate between exponential and Gaussian extremes that appear visually more like a Gamma distribution \citep{Houssais2012, Liu2019}.
The \citet{Houssais2012} experiments involved a binomal distribution of glass beads with diameters $0.7$mm and $2.2$mm in turbulent and supercritical flow conditions, while the \citet{Liu2019} experiments used sand having median diameter $1.1$mm. Flows were again turbulent and subcritical.
From this experimental record, the velocity distribution shape apparently does not consistently relate to super or sub-critical flow conditions, whether sediment grains are natural (sand, gravel) or synthetic (glass beads), or whether the flow is laminar or turbulent.
However there is a loose trend within the particle size. Smaller particles tend to show more exponential-like distributions \citep[e.g.][]{Fathel2015}, while larger ones show Gaussian distributions \citep[e.g.][]{Heyman2016}. This summary suggests that the shape of streamwise bed load velocity distribution could be controlled by the particle size.

@@

Several stochastic models have incorporated fluctuating driving and resisting terms into the Newtonian dynamics of individual grains to develop Langevin-like descriptions of bed load particle motions (sec. \ref{sec:langevin}). \citet{Fan2014} represented turbulent drag as Gaussian white noise and included a static Coulomb friction term to model particle-bed collisions, while \citet{Ancey2014} applied an Ornstein-Uhlenbeck process which combines the fluid and driving forces were lumped into a velocity-dependent force, while Gaussian white noise represents fluctuations.
Each of these models derives one end-member distribution from among the range of distributions reported in experiments.
A physical description for the full range of experimentally-observed bedload velocity distributions remains an elusive target.

@@

Particles respond to turbulent drag forces $F_D(t)$ and episodic particle-bed collision forces. In contrast to the computational physics approach, I do not aim to characterize the exact timeseries of the forces on an individual particle. Instead, I model the ensemble of possible force timeseries that particles could conceivably experience. Each possibility implies a different possible velocity timeseries $u(t)$ in the downstream direction.
The objective is to calculate the probability distribution $P(u)$ of this downstream velocity by averaging over the ensemble of forces.
I include the most realistic article-bed collision and fluid forces possible while still allowing for analytical solutions.

@@

\subsection{Turbulent fluid forces}
Fluid forces on a coarse particle in a viscous flow depend on the Reynolds number $\Re_p = d V/\nu$ defined by the particle size $d$, slip velocity $V$ between particle and fluid, and kinematic viscosity $\nu$.
These forces have been calculated analytically from the Navier-Stokes equations for vanishing $\Re_p$ and include acceleration, history, and velocity-dependent drag terms \citep{Hjelmfelt1966, Maxey1983, Auton1987}.
At realistic $\Re_p$, analytical results are inaccessible, so it is standard to include empirical corrections to the small $\Re_p$ formulas \citep{Schmeeckle2007,Clift1978}.
A dominant contribution to the downstream drag force $F_D$ on nearly spherical particles at large $\Re_p$ can be written $F_D = \frac{\pi}{8}
\rho_f d^2 C_D(\Re_p) |V|V$, where $\rho_f$ is the fluid density, $d$ is the particle diameter, $C_D(\Re_p)$ is an empirical drag coefficient, and $V = U-u$ is the slip velocity between the fluid ($U$) and particle ($u$) velocities \citep{Coleman1967, Schmeeckle2007, Dwivedi2012}.
In this study, I leave out acceleration and history terms for analytical tractability, although they are certainly relevant in coarse sediment transport \citep{Michaelides1997,Armenio2001}.

Drag forces have been argued to fluctuate rapidly compared to the inertial response times of submerged grains \citep{Fan2014}, and the magnitude of drag fluctuations has been observed to follow a Gaussian distribution \citep{Hofland2006,Schmeeckle2007,Dwivedi2010,Celik2014}.
These ideas allow two major simplifications of the above drag force.
First, the drag can be split into quasi-steady and fluctuating components \citep{Michaelides1997}, and second, drag fluctuations can be modelled a Gaussian white noise, where the fluctuations are symmetric, the autocorrelation time is negligible, and the strength of fluctuations is described by a diffusivity $D$ \citep{Fan2014,Ancey2014}. 
Defining $\bar{V}$ as a representative slip velocity to be specified more carefully later,  $\bar{C}_D$ as the empirical drag coefficient evaluated at this slip velocity, and $\xi(t)$ as a Gaussian white noise of mean $0$ and variance $1$ \citep{Gardiner1983}, the fluid drag can be written
\be F_D(t) = \Gamma + \sqrt{2 D } \eta(t), \label{eq:drag}\ee
where $\Gamma = \frac{\pi}{8}
\rho_f d^2 \bar{C}_D \bar{V}^2$ is the steady component of the drag.

@@

With the above forces, the Langevin equation $m\dot{u}(t) = F_D(t) + F_C(t)$ for the sediment dynamics becomes
\be m \dot{u}(t) = \Gamma + \sqrt{2D}\eta(t) - m u(t) \xi_{\nu, \ve}(t). \label{eq:langevin} \ee
This equation replaces the steady friction terms of earlier models with an episodic term, designed to provide a more realistic representation of particle-bed collisions.
Mathematically, equation \ref{eq:langevin} is a Langevin-like equation representing a jump-diffusion process \citep{Daly2006} with multiplicative Poisson noise \citep{Dubkov2016,Denisov2009}. 
Collisions introduce ``jumps" in velocity while turbulent generates "diffusion". The collision term is "multiplicative" in the sense that $u$ multiplies the Poisson noise.

Equations like \ref{eq:langevin} have long been studied in the stochastic physics literature \citep{Hanggi1978,VanDenBroeck1983}, but solving such equations remains extremely challenging \citep{Daly2010,Mau2014,Dubkov2019}.
One issue is that multiplicative white noises imply the prescription dilemma of stochastic calculus \citep{Risken1989,Gardiner1983}, meaning \ref{eq:langevin} is not defined without further specifying an integration rule \citep{Suweis2011}.
Here, the Ito interpretation (lower endpoint integration rule) is the physical choice the energy dissipated by collisions depends strictly on pre-collisional velocities, not post-collisional \citep[e.g.][]{Gardiner1983} since.
Given this integration rule, the remaining issues are to obtain the master equation characterizing the ensemble of velocities defined by \ref{eq:langevin}, and then to solve this equation for the velocity distribution $P(u)$.

\subsection{Chapman-Komogorov equation and particle-bed collision integral}

The equation governing the streamwise velocity distribution $P(u,t)$ is derived in appendix section \ref{sec:langmasterderiv} with a limiting procedure, providing
\be \nu^{-1}\partial_t P(u,t) = - \tilde{\Gamma} \partial_u P(u,t) + \tilde{D} \partial_u^2 P(u,t) + I_c(u,t). \label{eq:master} \ee
In this equation, I introduced the scaled parameters $\tilde{\Gamma} = \Gamma/(\nu m)$ and $\tilde{D} = D/(\nu m)$ that scale the steady and fluctuating components of the fluid force against the collision rate $\nu$. The term
\be \mathcal{I}_c(u,t) = - P(u,t) + \int_0^1 \frac{d\ve}{\ve}P\big(\frac{u}{\ve},t \big) \rho(\ve) \label{eq:colint} \ee
is a ``collision integral" term representing particle-bed collisions.

Equation \ref{eq:master} is a nonlocal extension of the Fokker-Planck equation used in earlier bed load models \citep{Fan2014,Ancey2014}. 
Nonlocality is introduced by the collision integral \ref{eq:colint} which transfers probability from higher pre-collisional velocities $u/\varepsilon$ to lower post-collisional velocities $u$.
This term is analogous to the collision integral within the Boltzmann equation of kinetic theory \citep{Duderstadt1979, Brilliantov2004}. Physically, it corresponds to binary collisions between particles having different masses and random resitution coefficients \cite{Serero2015} in the limit that the mass of one particle (here, the particle resting on the bed) diverges to infinity.
Mathematically, the collision integral represents the probability distribution of the product between the elasticity $\varepsilon$ and the downstream momentum $m u$ \citep[c.f.][]{Feller1968}.

Owing to its nonlocality, equation \ref{eq:master} does not admit analytical solutions as is, so further approximation is necessary.
Assuming the distribution of elasticity $\rho(\varepsilon)$ is sharply peaked at some most common (mode) value $\varepsilon'$, which is the case in experiments of rigid body collisions \citep{Glielmo2014}, allows for a Kramers-Moyal type expansion of the particle-bed collision integral \citep{Gardiner1983}.
Expanding all terms in the integrand except $\rho(\varepsilon)$ provides
\be \mathcal{I}_c(u,t) = -P(u,t) + \frac{1}{\varepsilon'}P\big(\frac{u}{\ve'},t \big) + \sum_{k=1}^\infty \frac{\alpha_k }{k!}(\ve - \ve')^k \Big[\frac{1}{\ve}P\big(\frac{u}{\ve}\big)\Big]^{(k)}\Big|_{\ve=\ve'},\label{eq:expansion}\ee
where the $\alpha_k = \int_0^1 d\ve \rho(\ve) (\ve-\ve')^k $ are the central moments of $\ve$ around the mode elasticity $\ve'$ and the superscript $(k)$ denotes the $k$th derivative.

@@

\subsection{Derivation of the bedload velocity distribution}
\label{sec:langsolution}
Hereafter I drop the prime on the most common streamwise restitution coefficient $\ve'$.
With the truncation to two terms, equation \ref{eq:master} gives 
\be 0 = -\tilde{\Gamma}\partial_u P(u) + \tilde{D}\partial_u^2 P(u) -P(u) + \frac{1}{\ve} P\big(\frac{u}{\ve}\big),\label{eq:governer} \ee
which is an advanced functional differential equation. This equation is ``functional" in the sense that it is nonlocal in velocity due to the last term on the right hand side, and it is ``advanced" in that $u/\ve$ is advanced beyond the argument $u$ involved in the rest of this equation (since $0<\ve<1$).
Such equations seen some attention in the mathematics literature, where they are sometimes called pantograph equations \citep{Hall1989, Kim1998,Zaidi2015}.

In the appendix section \ref{sec:langsteadyderiv} I solve equation \ref{eq:governer} using Laplace transforms, providing
\begin{multline} P(u) = \frac{\theta(-u)}{K_+}\sum_{l=0}^\infty \frac{\ve^{-l}e^{\lambda_+\ve^{-l}u}}{\prod_{m=1}^l(-\tilde{D} \lambda_+^2 \ve^{-2m} + \tilde{\Gamma} \lambda_+ \ve^{-m} + 1) } 
	\\+ \frac{\theta(u)}{K_-}\sum_{l=0}^\infty \frac{\ve^{-l}e^{\lambda_-\ve^{-l}u}}{\prod_{m=1}^l(- \tilde{D} \lambda_-^2 \ve^{-2m} + \tilde{\Gamma} \lambda_- \ve^{-m} + 1) } \label{eq:steadystate}. \end{multline}
The factors $\lambda_\pm$ are defined in appendix equations \ref{eq:lambdas}. They are proportional to $\tilde{\Gamma}/\tilde{D}$. 
The normalization factors are $K_\pm$ are 
\be K_\pm = \tilde{D}(\lambda_+-\lambda_-)\prod_{l=1}^\infty (- \tilde{D}\lambda_\pm^2 \ve^{2l} +\tilde{\Gamma} \lambda_\pm \ve^{l} + 1). \ee
Although this velocity distribution has a complex mathematical structure, one can verify that this is a normalized probability distribution ($\int du P(u) = 1$) which has very simple limiting behaviors as the mode elasticity $\ve$ approaches fully elastic ($\ve=1$) and inelastic ($\ve = 0$) values. The complexity of this result is not surprising given how few analytical solutions are available in granular gas theories with similar nonlocal collision integrals \citep[e.g.][]{Brilliantov2004}.

@@

In fact, the apparent transition from exponential to Gaussian in figure \ref{fig:fig2} can be demonstrated from equation \ref{eq:steadystate}. Despite its complex appearance, simple Gaussian and exponential forms derive from this equation as exact mathematical limits.
When particle-bed collisions are completely inelastic ($\ne = 0$), \ref{eq:steadystate} becomes an exponential distribution, and when they are completely elastic ($\nu = 1$), \ref{eq:steadystate} becomes Gaussian.
Figure \ref{fig:fig3} demonstrates in detail the approach of the distribution toward these limits.
\begin{figure}
	\centerline{\includegraphics{./figures/ch5/Fig3asymptotic.png}}
	\caption{The particle velocity distribution approaches an exponential distribution in (a) as particle-bed collisions become extremely elastic ($\ve \rightarrow 1$), and it approaches a Gaussian in (b) as they become extremely inelastic ($\ve \rightarrow 0$). On the abscissa, the mean sediment velocity is standardized by its mean $\bar{u}$ and standard deviation $\sigma_u$. }
	\label{fig:fig3}
\end{figure}

The exponential limit of \ref{eq:steadystate} as $\ve \rightarrow 0$ is easy to see. Taking $\ve \rightarrow 0 $ in \ref{eq:steadystate}, all terms in the series except for that with $l=0$ become exponentially small, leaving behind the same two-sided exponential distribution derived by \cite{Fan2014} up to differences in notation:
\be P(u) = \frac{d}{\sqrt{\gamma^2 + 4d}}e^{\frac{\gamma u }{2 d} - \frac{\sqrt{\gamma^2 + 4 d}|u|}{2d}}. \ee
Thus, for bed load transport conditions with typically very inelastic particle-bed collisions, we can expect exponential-like velocities and large deviations from a Gaussian behavior.

@@

Although in reality, the turbulent forces on moving sediment particles vary in a complex spatio-temporal way, I have approximated the fluid forces on bed load particles as spatially uniform Gaussian white noise. Even though the non-Gaussian and history-dependent aspects of fluid turbulence certainly do impact sediment entrainment \citep{Cameron2020,Celik2014}, this flow model appears more or less justified since sediment transport experiments provide similar velocity distributions regardless of whether the flow is viscous or turbulent \citep{Lajeunesse2010, Charru2004}, and since particle relaxation times are typically long compared to the timescales of turbulent fluctuations \citep{Hofland2006,Schmeeckle2007,Nakagawa1981}.

The model developed in this chapter described particle-bed collision forces as a sequence of instantaneous impulses in an approach that is reminiscent of the kinetic theory of gases \citep{Landau1969}.
The effect of each collision on the streamwise particle velocities was described by a simple elasticity-like coefficient.
Although such approximate descriptions of particle-particle collisions are common in the theory of granular gases, the elasticity coefficient introduced here is not equivalent to the ``coefficient of restitution" typically applied in granular gas theory.
The coefficient of resitution is defined normal to the contact plane of two particles during a collision, and it characterizes energy loss due to deformation of the colliding particles in the absence of a viscous fluid \citep{Brach1992,Ismail2008}.

In bedload transport, colliding particles are submerged in a viscous fluid, and this introduces additional damping processes besides particle deformation \citep{Joseph2001,Yang2006,Schmeeckle2001}.
The model developed here did not consider these viscous forces, nor did it explicitly model the geometry of collisions, as $\ve$ was defined as a parameter applying to the downstream velocity only.
Thus, although I have included episodic particle-bed collisions in a stochastic sediment transport model for the first time, the key parameter ($\ve$) in this formulation has a rather heuristic character and is not clearly related to the coefficient of restitution from granular gas theory.
Future studies should formulate particle-bed collisions in stochastic sediment transport models considering more details of collision geometry \citep{Sekine1992}, fluid-particle interactions \citep{Marshall2011}, and traditional restitution \citep{Brach1989} using the present study as a starting point.

@@

An alternative viewpoint, formulated in this chapter, is that the shape of the velocity distribution originates from granular interactions.
Particle size enters the Langevin model \ref{eq:langevin} explicitly within the steady component of the drag, but it may also enter implicitly through the elasticity parameter $\ve$.
The analytical velocity distribution \ref{eq:steadystate} was fit to six different experimental datasets in figure \ref{fig:fig4ch5}, and the best fit parameters required to for these fits were presented in table \ref{tab:calib}.
These fit parameters suggest collisions are generally less elastic for smaller particle sizes, whereas they are more elastic for larger particle sizes. This suggests that the amount of momentum dissipated by collisions may depend on particle size.

@@

\section{Summary}
\label{sec:langconclusion}
This chapter has presented a Langevin description of individual bedload particles saltating downstream through episodic collisions.
The model suggests that episodic particle-bed collisions control the shape of the particle velocity distribution, and it is the first model to describe the full range of bedload velocity distributions that have been reported in experimental studies.

>>>>

High-speed video experiments have measured different streamwise particle velocity distributions without indicating why one distribution or another appears in a given set of hydraulic and sedimentary conditions.
One set of studies has shown exponential particle velocity distributions \citep{Charru2004,Lajeunesse2010,Roseberry2012,Seizilles2014,Fathel2015,Fathel2016}.
These experiments involve uniformly-sized small sands or glass beads ($~0.05-2$mm) in turbulent and subcritical flows, but not always; the experiments of \citet{Lajeunesse2010} were supercritical, while the experiments of \citet{Charru2004} and \citet{Seizilles2014} were viscous.
A second set of studies show Gaussian particle velocity distributions \citep{Ancey2014,Heyman2016,Martin2012}. In these experiments, particles are typically larger ($2-8$mm) uniformly-sized gravels or glass beads, and flows are generally turbulent and super-critical.
Other experiments display velocity distributions that are intermediate between exponential and Gaussian extremes that appear visually more like a Gamma distribution \citep{Houssais2012, Liu2019}.
The \citet{Houssais2012} experiments involved a binomal distribution of glass beads with diameters $0.7$mm and $2.2$mm in turbulent and supercritical flow conditions, while the \citet{Liu2019} experiments used sand having median diameter $1.1$mm. Flows were again turbulent and subcritical.
From this experimental record, the velocity distribution shape apparently does not consistently relate to super or sub-critical flow conditions, whether sediment grains are natural (sand, gravel) or synthetic (glass beads), or whether the flow is laminar or turbulent.
However there is a loose trend within the particle size. Smaller particles tend to show more exponential-like distributions \citep[e.g.][]{Fathel2015}, while larger ones show Gaussian distributions \citep[e.g.][]{Heyman2016}. This summary suggests that the shape of streamwise bed load velocity distribution could be controlled by the particle size.

@@

Several stochastic models have incorporated fluctuating driving and resisting terms into the Newtonian dynamics of individual grains to develop Langevin-like descriptions of bed load particle motions (sec. \ref{sec:langevin}). \citet{Fan2014} represented turbulent drag as Gaussian white noise and included a static Coulomb friction term to model particle-bed collisions, while \citet{Ancey2014} applied an Ornstein-Uhlenbeck process which combines the fluid and driving forces into a simplified velocity-dependent force, while Gaussian white noise represents fluctuations.
Each of these models derives one end-member distribution from among the range of distributions reported in experiments.
A physical description for the full range of experimentally-observed bedload velocity distributions remains an elusive target.

@@

Particles respond to turbulent drag forces $F_D(t)$ and episodic particle-bed collision forces. In contrast to the computational physics approach, I do not aim to characterize the exact time series of the forces on an individual particle. Instead, I model the ensemble of possible force time series that particles could conceivably experience. Each force time series realization implies a different velocity time series $u(t)$ in the downstream direction.
The objective is to calculate the probability distribution $P(u)$ of this downstream velocity by averaging over the ensemble of forces.
I include the most realistic article-bed collision and fluid forces possible while still allowing for analytical solutions.

@@

\subsection{Turbulent fluid forces}
Fluid forces on a coarse particle in a viscous flow depend on the Reynolds number $\Rey_p = d V/\lambda$ defined by the particle size $d$, slip velocity $V$ between particle and fluid, and kinematic viscosity $\lambda$.
These forces have been calculated analytically from the Navier-Stokes equations for vanishing $\Rey_p$ and include acceleration, history, and velocity-dependent drag terms \citep{Hjelmfelt1966, Maxey1983, Auton1987}.
At realistic $\Rey_p$, analytical results are inaccessible, so it is standard to include empirical corrections to the small $\Rey_p$ formulas \citep{Schmeeckle2007,Clift1978}.
A dominant contribution to the downstream drag force $F_D$ on nearly spherical particles at large $\Rey_p$ can be written $F_D = \frac{\pi}{8}
\rho_f d^2 C_D(\Re_p) |V|V$, where $\rho_f$ is the fluid density, $d$ is the particle diameter, $C_D(\Rey_p)$ is an empirical drag coefficient, and $V = U-u$ is the slip velocity between the fluid ($U$) and particle ($u$) velocities \citep{Coleman1967, Schmeeckle2007, Dwivedi2012}.
In this study, I leave out acceleration and history terms for analytical tractability, although they are certainly relevant in coarse sediment transport \citep{Michaelides1997,Armenio2001}.

Drag forces have been argued to fluctuate rapidly compared to the inertial response times of submerged grains \citep{Fan2014}, and the magnitude of drag fluctuations has been observed to follow a Gaussian distribution \citep{Hofland2006,Schmeeckle2007,Dwivedi2010,Celik2014}.
These ideas allow two major simplifications of the above drag force.
First, the drag can be split into quasi-steady and fluctuating components \citep{Michaelides1997}, and second, drag fluctuations can be modelled as a Gaussian white noise, where the fluctuations are symmetric, the autocorrelation time is negligible, and the strength of fluctuations is characterized by a diffusivity $D$ \citep{Fan2014,Ancey2014}. 
Defining $\bar{V}$ as a representative slip velocity to be specified more carefully later,  $\bar{C}_D$ as the empirical drag coefficient evaluated at this slip velocity, and $\xi(t)$ as a Gaussian white noise of mean $0$ and variance $1$ \citep{Gardiner1983}, the fluid drag can be written
\be F_D(t) = \Gamma + \sqrt{2 D } \eta(t), \label{eq:drag}\ee
where $\Gamma = \frac{\pi}{8}
\rho_f d^2 \bar{C}_D \bar{V}^2$ is the steady component of the drag.

@@

With the above forces, the Langevin equation $m\dot{u}(t) = F_D(t) + F_C(t)$ for the sediment dynamics becomes
\be m \dot{u}(t) = \Gamma + \sqrt{2D}\eta(t) - m u(t) \xi_{\nu, \ve}(t). \label{eq:langevin} \ee
This equation replaces the steady friction terms of earlier models with an episodic term, designed to provide a more realistic representation of particle-bed collisions.
Mathematically, equation \ref{eq:langevin} is a Langevin-like equation representing a jump-diffusion process \citep{Daly2006} with multiplicative Poisson noise \citep{Dubkov2016,Denisov2009}. 
Collisions introduce ``jumps" in velocity while turbulence generates "diffusion". The collision term is "multiplicative" in the sense that $u$ multiplies the Poisson noise.

Equations like \ref{eq:langevin} have long been studied in the stochastic physics literature \citep{Hanggi1978,VanDenBroeck1983}, but solving such equations remains extremely challenging \citep{Daly2010,Mau2014,Dubkov2019}.
One issue is that multiplicative white noises imply the prescription dilemma of stochastic calculus \citep{Risken1989,Gardiner1983}, meaning \ref{eq:langevin} is not defined without further specifying an integration rule \citep{Suweis2011}.
Here, the Ito interpretation (lower endpoint integration rule) is the physical choice the energy dissipated by collisions depends strictly on pre-collisional velocities, not post-collisional \citep[e.g.][]{Gardiner1983}.
Given this integration rule, the remaining issues are to obtain the master equation characterizing the ensemble of velocities defined by \ref{eq:langevin}, and then to solve this equation for the velocity distribution $P(u)$.

\subsection{Chapman-Kolmogorov equation and particle-bed collision integral}

The equation governing the streamwise velocity distribution $P(u,t)$ is derived in appendix section \ref{sec:langmasterderiv} with a limiting procedure, providing
\be \nu^{-1}\partial_t P(u,t) = - \tilde{\Gamma} \partial_u P(u,t) + \tilde{D} \partial_u^2 P(u,t) + I_c(u,t). \label{eq:master} \ee
In this equation, I introduced the scaled parameters $\tilde{\Gamma} = \Gamma/(\nu m)$ and $\tilde{D} = D/(\nu m)$ that scale the steady and fluctuating components of the fluid force against the collision rate $\nu$. The term
\be I_c(u,t) = - P(u,t) + \int_0^1 \frac{d\ve}{\ve}P\big(\frac{u}{\ve},t \big) \rho(\ve) \label{eq:colint} \ee
is a ``collision integral" term representing particle-bed collisions.

Equation \ref{eq:master} is a nonlocal extension of the Fokker-Planck equation used in earlier bed load models \citep{Fan2014,Ancey2014}. 
Nonlocality is introduced by the collision integral \ref{eq:colint} which transfers probability from higher pre-collisional velocities $u/\varepsilon$ to lower post-collisional velocities $u$.
This term is analogous to the collision integral within the Boltzmann equation of kinetic theory \citep{Duderstadt1979, Brilliantov2004}. Physically, it corresponds to binary collisions between particles having different masses and random restitution coefficients \cite{Serero2015} in the limit that the mass of one particle (here, the particle resting on the bed) diverges to infinity.
Mathematically, the collision integral represents the probability distribution of the product between the elasticity $\varepsilon$ and the downstream momentum $(m u)$ \citep[cf.][]{Feller1968}.

Owing to its nonlocality, equation \ref{eq:master} does not admit analytical solutions as is, so further approximation is necessary.
Assuming the distribution of elasticity $\rho(\varepsilon)$ is sharply peaked at some most common (mode) value $\varepsilon'$, which is the case in experiments of rigid body collisions \citep{Glielmo2014}, allows for a Kramers-Moyal type expansion of the particle-bed collision integral \citep{Gardiner1983}.
Expanding all terms in the integrand except $\rho(\varepsilon)$ provides
\be \mathcal{I}_c(u,t) = -P(u,t) + \frac{1}{\varepsilon'}P\big(\frac{u}{\ve'},t \big) + \sum_{k=1}^\infty \frac{\alpha_k }{k!}(\ve - \ve')^k \Big[\frac{1}{\ve}P\big(\frac{u}{\ve}\big)\Big]^{(k)}\Big|_{\ve=\ve'},\label{eq:expansion}\ee
where the $\alpha_k = \int_0^1 d\ve \rho(\ve) (\ve-\ve')^k $ are the central moments of $\ve$ around the modal elasticity $\ve'$ and the superscript $(k)$ denotes the $k$th derivative.

@@

\subsection{Derivation of the bedload velocity distribution}
\label{sec:langsolution}
Hereafter I drop the prime on the most common streamwise restitution coefficient $\ve'$.
With the truncation to two terms, equation \ref{eq:master} gives 
\be 0 = -\tilde{\Gamma}\partial_u P(u) + \tilde{D}\partial_u^2 P(u) -P(u) + \frac{1}{\ve} P\big(\frac{u}{\ve}\big),\label{eq:governer} \ee
which is an advanced functional differential equation. This equation is ``functional" in the sense that it is nonlocal in velocity due to the last term on the right hand side, and it is ``advanced" in that $u/\ve$ is advanced beyond the argument $u$ involved in the rest of this equation (since $0<\ve<1$).
Such equations have seen some attention in the mathematics literature, where they are sometimes called pantograph equations \citep{Hall1989, Kim1998,Zaidi2015}.

In the appendix section \ref{sec:langsteadyderiv} I solve equation \ref{eq:governer} using Laplace transforms, providing
\begin{multline} P(u) = \frac{\theta(-u)}{K_+}\sum_{l=0}^\infty \frac{\ve^{-l}e^{\lambda_+\ve^{-l}u}}{\prod_{m=1}^l(-\tilde{D} \lambda_+^2 \ve^{-2m} + \tilde{\Gamma} \lambda_+ \ve^{-m} + 1) } 
	\\+ \frac{\theta(u)}{K_-}\sum_{l=0}^\infty \frac{\ve^{-l}e^{\lambda_-\ve^{-l}u}}{\prod_{m=1}^l(- \tilde{D} \lambda_-^2 \ve^{-2m} + \tilde{\Gamma} \lambda_- \ve^{-m} + 1) } \label{eq:steadystate}. \end{multline}
The factors $\lambda_\pm$ are defined in appendix equations \ref{eq:lambdas}. They are proportional to $\tilde{\Gamma}/\tilde{D}$. 
The normalization factors $K_\pm$ are 
\be K_\pm = \tilde{D}(\lambda_+-\lambda_-)\prod_{l=1}^\infty (- \tilde{D}\lambda_\pm^2 \ve^{2l} +\tilde{\Gamma} \lambda_\pm \ve^{l} + 1). \ee
Although this velocity distribution has a complex mathematical structure, one can verify that this is a normalized probability distribution ($\int du P(u) = 1$) which has very simple limiting behaviors as the mode elasticity $\ve$ approaches fully elastic ($\ve=1$) and inelastic ($\ve = 0$) values. The complexity of this result is not surprising given how few analytical solutions are available in granular gas theories with similar nonlocal collision integrals \citep[e.g.][]{Brilliantov2004}.

@@

In fact, the apparent transition from exponential to Gaussian in figure \ref{fig:fig2} can be demonstrated from equation \ref{eq:steadystate}. Despite its complex appearance, simple Gaussian and exponential forms derive from this equation as exact mathematical limits.
When particle-bed collisions are completely inelastic ($\ve = 0$), \ref{eq:steadystate} becomes an exponential distribution, and when they are completely elastic ($\ve = 1$), \ref{eq:steadystate} becomes Gaussian.
Figure \ref{fig:fig3} demonstrates in detail the approach of the distribution toward these limits.
\begin{figure}
	\centerline{\includegraphics{./figures/ch5/Fig3asymptotic.png}}
	\caption{The particle velocity distribution approaches an exponential distribution in (a) as particle-bed collisions become extremely elastic ($\ve \rightarrow 1$), and it approaches a Gaussian in (b) as they become extremely inelastic ($\ve \rightarrow 0$). On the abscissa, the mean sediment velocity is standardized by its mean $\bar{u}$ and standard deviation $\sigma_u$. }
	\label{fig:fig3}
\end{figure}

The exponential limit of \ref{eq:steadystate} as $\ve \rightarrow 0$ is easy to see. Taking $\ve \rightarrow 0 $ in \ref{eq:steadystate}, all terms in the series except for the one at $l=0$ become exponentially small, leaving behind the same two-sided exponential distribution derived by \cite{Fan2014} up to differences in notation:
\be P(u) = \frac{d}{\sqrt{\gamma^2 + 4d}}e^{\frac{\gamma u }{2 d} - \frac{\sqrt{\gamma^2 + 4 d}|u|}{2d}}. \ee
Thus, for bed load transport conditions with typically very inelastic particle-bed collisions, we can expect exponential-like velocities and large deviations from a Gaussian behavior.

@@

Although in reality, the turbulent forces on moving sediment particles vary in a complex spatio-temporal way, I have approximated the fluid forces on bed load particles as spatially uniform Gaussian white noise. Even though the non-Gaussian and history-dependent aspects of fluid turbulence certainly do impact sediment entrainment \citep{Cameron2020,Celik2014}, this flow model appears more or less justified since sediment transport experiments provide similar velocity distributions regardless of whether the flow is viscous or turbulent \citep{Lajeunesse2010, Charru2004}, and since particle relaxation times are typically long compared to the time scales of turbulent fluctuations \citep{Hofland2006,Schmeeckle2007,Nakagawa1981}.

The model developed in this chapter described particle-bed collision forces as a sequence of instantaneous impulses in an approach that is reminiscent of the kinetic theory of gases \citep{Landau1969}.
The effect of each collision on the streamwise particle velocity was described by a simple elasticity-like coefficient.
Although such approximate descriptions of particle-particle collisions are common in the theory of granular gases, the elasticity coefficient introduced here is not equivalent to the ``coefficient of restitution" typically applied in granular gas theory.
The coefficient of restitution is defined normal to the contact plane of two particles during a collision, and it characterizes energy loss due to deformation of the colliding particles in the absence of a viscous fluid \citep{Brach1992,Ismail2008}.

In bedload transport, colliding particles are submerged in a viscous fluid, and this introduces additional damping processes besides particle deformation \citep{Joseph2001,Yang2006,Schmeeckle2001}.
The model developed here does not consider these viscous forces, nor does it explicitly model the geometry of collisions, as $\ve$ was defined as a parameter applying to the downstream velocity only.
Thus, although I have included episodic particle-bed collisions in a stochastic sediment transport model for the first time, the key parameter ($\ve$) in this formulation has a rather heuristic character and is not clearly related to the coefficient of restitution from granular gas theory.
Future studies should formulate particle-bed collisions in stochastic sediment transport models considering more details of collision geometry \citep{Sekine1992}, fluid-particle interactions \citep{Marshall2011}, and traditional restitution \citep{Brach1989} using the present study as a starting point.

@@

An alternative viewpoint, formulated in this chapter, is that the shape of the velocity distribution originates from granular interactions.
Particle size enters the Langevin model \ref{eq:langevin} explicitly within the steady component of the drag, but it may also enter implicitly through the elasticity parameter $\ve$.
The analytical velocity distribution \ref{eq:steadystate} was fit to six different experimental datasets in figure \ref{fig:fig4ch5}, and the best fit parameters were presented in table \ref{tab:calib}.
These fit parameters suggest collisions are generally less elastic for smaller particle sizes, whereas they are more elastic for larger particle sizes. This suggests that the amount of momentum dissipated by collisions may depend on particle size.

@@

\section{Summary}
\label{sec:langconclusion}
This chapter has presented a Langevin description of individual bedload particles saltating downstream through episodic collisions.
The model suggests that episodic particle-bed collisions control the shape of the particle velocity distribution, and it is the first model to describe the full range of bedload velocity distributions reported in experimental studies.
